Minimax-optimal Laplacian Eigenmaps regression over Sobolev Spaces with Neighborhood Graphs

### 2021-03-17 manuscripts/laplacian_eigenmaps/laplacian_eigenmaps_regression.tex

\documentclass{article}
\usepackage{amsmath}
\usepackage{amsfonts, amsthm, amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\hypersetup{
	colorlinks=true,
	linkcolor=blue,
	citecolor=blue
}
\usepackage[parfill]{parskip}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{enumerate}
\usepackage[shortlabels]{enumitem}
\usepackage{fullpage}
\usepackage{mathtools}
\usepackage{tikz}

\usepackage{natbib}
\renewcommand{\bibname}{REFERENCES}
\renewcommand{\bibsection}{\subsubsection*{\bibname}}

@@

\newcommand{\emF}{\mathbb{F}_n}
\newcommand{\emG}{\mathbb{G}_n}
\newcommand{\emP}{\mathbb{P}_n}
\newcommand{\F}{\mathcal{F}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\Rd}{\Reals^d}
\newcommand{\Nbb}{\mathbb{N}}

@@

\textbf{(1) Geometric graphs.} In graph-based learning, one observes data $X_1,\ldots,X_n$ sampled independently from an unknown distribution $P$, and forms a geometric graph $G$---with edges $e_{ij}$ corresponding to proximity between samples $X_i$ and $X_j$---over the observed data. Geometric graphs encode information about $P$ in an extremely general manner, and can be leveraged to conduct many different fundamental statistical tasks. These include clustering, semi-supervised learning, classification, and regression---both estimation and goodness-of-fit testing. Though much theoretical work has been done on the consistency of graph-based learning methods---and more recently, rates of convergence have been established for some problems---little is known so far about their optimality, even for classic statistical tasks.

\textbf{(2) What we study: Regression and Laplacian eigenmaps.} In this paper we focus on regression where in addition to the design points $X_1,\ldots,X_n$ one observes real-valued responses $Y_1,\ldots,Y_n$, and seeks to learn the unknown regression function $f_0(x) := E[Y|X = x]$. We consider both the estimation and goodness-of-fit testing problems. The methods we will study are based on \emph{Laplacian eigenmaps}, first introduced by \cite{belkin03a}, which projects the response vector $Y = (Y_1,\ldots,Y_n)$ onto the span of the leading eigenvectors of the graph Laplacian $L$. The Laplacian $L$ is a difference operator, acting on functions $f: \{X_1,\ldots,X_n\} \to \Reals$ as follows,
\begin{equation}
\label{eqn:graph_laplacian}
Lf(X_i) = \sum_{j = 1}^{n} \bigl(f(X_i) - f(X_j)\bigr)e_{ij}. 
\end{equation}
The Laplacian $L$ can be seen as a discretization of the Laplace-Beltrami operator $\Delta_P$ \textcolor{red}{(...)}. The eigenvectors of $L$ form an orthonormal basis of $L^2(X)$, and are approximations to the eigenvectors of $\Delta_P$. Their corresponding eigenvalues are estimates of~\textcolor{red}{(...)}, and provide a notion of smoothness for each eigenvector---the smaller the eigenvalue, the smoother the corresponding eigenvector.

\textbf{(3) Why we study.} Laplacian eigenmaps is thus an example of a \emph{spectral series} estimator. A spectral series estimator is a special case of an orthogonal series estimator, one of the most classical methods of non-parametric regression. Traditionally, orthogonal series estimators are defined with respect to a fixed set of basis functions, orthogonalized with respect to some pre-determined reference measure $Q$. Spectral series estimators take a particular orthogonal basis, the eigenfunctions of $\Delta_Q$. In contrast, eigenvectors of the graph Laplacian are data-dependent objects, and adapt to the geometry of the design distribution $P$ in a rich manner.

The theoretical properties of (classical) spectral series estimators are by this point well-understood \textcolor{red}{(references)}. In particular, they are minimax optimal for nonparametric regression over H\"{o}lder and Sobolev spaces. However, in practice these estimators suffer from some serious drawbacks. Finding the eigenfunctions of $\Delta_Q$ is in general non-trivial. The statistical optimality of such estimators also hinges on the reference measure $Q$ being equal to $P$, an unrealistic assumption when $P$ is unknown. If $Q \neq P$, the standard recommendation is to instead do least squares---this can be statistically rate-optimal, but is numerically unstable when dimension $d$ is even moderately large.

\textbf{(4) Our contributions.} Laplacian eigenmaps avoid these drawbacks \textcolor{red}{(...)} On the other hand their statistical properties are not as well understood \textcolor{red}{(...)}. The primary contribution of our paper is to fill this theoretical gap, by answering the following question:

@@

Broadly speaking, we show that when the regression function $f_0$ is smooth, in the sense of having bounded derivatives in the Sobolev sense, Laplacian eigenmaps is statistically minimax optimal. This statement holds for different relations between the dimension $d$ and number of derivatives $s$, depending on the problem (estimation or testing). \textcolor{red}{(TODO): Add two tables summarizing your major results here.}

@@

\textcolor{red}{TODO}: Define $\Leb^2(\mc{X}), \Leb^2(P)$ and $\Leb^2(P_n)$.

For a square, symmetric matrix $A \in \Reals^{n \times n}$, an eigenvector $v$ of $A$ is a \textcolor{red}{unit-norm} solution to the eigenproblem
\begin{equation*}
Av = \lambda v,
\end{equation*}
and $\lambda$ is the corresponding eigenvalue.

\section{Setup, Background, and Overview of Results}
\label{sec:setup_main_results}

In this section, we begin by giving a precise definition of Laplacian eigenmaps. We then review minimax rates for non-parametric regression over Sobolev spaces, allowing us to summarize our main results.

\subsection{Non-parametric regression with Laplacian Eigenmaps}
\label{sec:regression_laplacian_eigenmaps}

\textbf{(1) Formal setup.}
We will operate in the usual setting of non-parametric regression with random design. We observe independent random samples $(X_1,Y_1),\ldots,(X_n,Y_n)$, where the design points $X_1,\ldots,X_N$ are drawn i.i.d from a distribution $P$ supported on a compact set $\mc{X} \subseteq \Rd$, and responses
\begin{equation}
\label{eqn:model}
Y_i = f_0(X_i) + \epsilon_i,
\end{equation}
with regression function $f_0: \mc{X} \to \Reals$, and $\epsilon_i \sim N(0,1)$ independent Gaussian noise. For simplicity we will assume throughout that the noise has unit-variance, but all of our results extend in a straightforward manner to the case where the variance is equal to a known positive value.

\textbf{(2) Laplacian eigenmaps.} Laplacian eigenmaps constructs an estimate of the regression function $f_0$ using the eigenvectors of a neighborhood graph Laplacian operator. For a positive, radially symmetric kernel $\eta: [0,\infty) \to [0,\infty)$, and a bandwidth parameter $\varepsilon > 0$, the \emph{neighborhood graph Laplacian} operator $L_{n,\varepsilon}$ is
\begin{equation}
\label{eqn:neighborhood_graph_laplacian}
L_{n,\varepsilon}u(x) = \frac{1}{n\varepsilon^{d + 2}} \sum_{i = 1}^{n} \bigl(u(x) - u(X_j)\bigr) \eta\biggl(\frac{\|x - X_j\|}{\varepsilon}\biggr)
\end{equation}
The constant scaling $(n\varepsilon^{d + 2})^{-1}$ is purely for convenience in taking limits as $n \to \infty, \varepsilon \to 0$. Also, although~\eqref{eqn:neighborhood_graph_laplacian} makes sense for any $x \in \mc{X}$ and $u: \mc{X} \to \Reals$, we will typically think of $x \in \{X_1,\ldots,X_n\}$ and $u = (u(X_1),\ldots,u(X_n)) \in \Leb^2({\bf X})$. The graph Laplacian is positive semi-definite, and we will index its eigenpairs $(\lambda_1,v_1),\ldots,(\lambda_n,v_n)$ in ascending order of eigenvalue, $0 = \lambda_1 \leq \ldots \leq \lambda_n$.

For a given $K \in \{1,\ldots,n\}$, the order-$K$ Laplacian eigenmaps estimator simply projects the response vector ${\bf Y}$ onto the first $K$ eigenvectors of $L_{n,\varepsilon}$: letting $V_K \in \Reals^{n \times K}$ be the matrix with columns $v_1,\ldots,v_K$, we have that
\begin{equation}
\label{eqn:laplacian_eigenmaps_estimator}
\wh{f} := \sum_{k = 1}^{K} \dotp{Y}{v_k}_{n} v_k = \frac{1}{n} V_K V_K^{\top} Y.
\end{equation} 
Since the eigenvectors $v_1,\ldots,v_n$ are orthogonalized with respect to the $L^2(P_n)$ inner product, the estimator $\wh{f}$ is the least-squares solution to the linear regression problem with responses $Y_1,\ldots,Y_n$ and features $v_1,\ldots,v_K$. Note that $\wh{f}$ is defined only in-sample, that is, only at the design points $X_1,\ldots,X_n$---we will consider the question of out-of-sample estimation later in Section~\ref{sec:out_of_sample}.

If $\wh{f}$ is a reasonable estimate of $f_0$, then the test statistic
\begin{equation}
\label{eqn:laplacian_eigenmaps_test}
\wh{T} := \|\wh{f}\|_n^2
\end{equation}
is in turn a reasonable estimate of $\|f_0\|_{P}^2$, and can be used to distinguish whether or not $f_0 = 0$.

\subsection{Sobolev Classes}
\label{sec:sobolev}
Before we give an overview of our main results, we will first pause to review the Sobolev classes and \textcolor{red}{associated} minimax rates.

\textbf{(1) Sobolev norms, semi-norms, and balls.} 
The Sobolev class $H^s(P)$ consists of all those functions $f \in \Leb^2(P)$ which have weak partial derivatives of all orders $j = 1,\ldots,s$, all of which are bounded in $\Leb^2(P)$ norm. The Sobolev norm $\|f\|_{H^s(P)}$ is defined according to
\begin{equation*}
\|f\|_{H^s(P)}^2 := \|f\|_{\Leb^2(P)}^2 + \sum_{j = 1}^{s} \|D^jf\|_{\Leb^2(P)}^2,
\end{equation*}
where $D^jf := \sum_{|\alpha| = j}D^{\alpha}f$ is the sum of all weak partial derivatives of order $j$, and we use the multiindex notation $D^{\alpha}f = \partial^j f/(\partial^{\alpha_1}x^1\ldots \partial^{\alpha_d}x^d)$ for multiindex $\alpha$. The Sobolev ball of radius $M$ consists of all those functions $f \in H^s(P)$ with Sobolev norm no greater than $M$,
\begin{equation*}
H^s(P;M) := \bigl\{f \in H^s(P): \|f\|_{H^s(P)} \leq M\bigr\}
\end{equation*}

\textcolor{red}{(TODO)}: 
\begin{itemize}
	\item Possibly define the iterated Laplace-Beltrami operator. Point out that for $H^{2s}(P)$ functions, the Sobolev semi-norm can be stated in terms of inner products using the iterated Laplace-Beltrami operator.
	\item Introduce the notation $H^s(\mc{X}) = H^s(\nu)$, where $\nu$ is Lebesgue measure.
\end{itemize}

\textbf{(2) Zero-trace and periodic conditions.}
When $s \geq 1$, our results regarding Laplacian eigenmaps will require that $f_0$ satisfy some boundary assumptions. In particular, we will assume that $f_0$ belongs to the set of zero trace Sobolev functions $H_0^{s}(P)$, meaning \textcolor{red}{roughly} that $f_0 \in H^s(P)$ additionally satisfies,
\begin{equation}
\label{eqn:zero_trace}
D^jf(x) = 0,~~\textrm{for all $x \in \partial \mc{X}$ and $j = 0,\ldots,s - 1$.}
\end{equation}
\textcolor{red}{(TODO)}: This is in spirit right but technically incoherent unless $f \in C^{\infty}(\mc{X})$. Need to use the formal definition of trace.

Let us justify why we impose a boundary condition such as zero-trace. For this part only we concentrate on the special case of $d = 1$ and the domain $\mc{X} = [0,1]$. In this case, there exists another way to define Sobolev space, via a sequence space representation (equiv Fourier transform). Let $\psi_1,\psi_2,\ldots$ denote the trigonometric basis of $\Leb^2([0,1])$,\footnote{To be explicit, $\psi_1(x) := 1$, $\psi_{2k}(x) := \sqrt{2} \cos(2\pi k x)$ and $\psi_{2k + 1}(x) := \sqrt{2} \sin(2\pi k x)$} and let $\Theta \subset \ell^2(\mathbb{N})$ be the Sobolev ellipsoid $\Theta = \{\theta \in \ell^2(\mathbb{N}): \sum_{k = 1}^{\infty} k^{2s} \theta_k^2 \leq M\}$. The function class $\{f = \sum_{k = 1}^{\infty} \theta_k \psi_k: \theta \in \Theta\}$ is \textcolor{red}{equivalent to} the \emph{periodic Sobolev space} $H_{\mathrm{per}}^s([0,1])$, defined as
\begin{equation}
\label{eqn:periodic_sobolev_space}
H_{\mathrm{per}}^s([0,1]) = \Bigl\{f \in H^s([0,1],M): D^jf(0) = D^jf(1)~~\textrm{for $j = 0,\ldots, s - 1$} \Bigr\}
\end{equation}
It is known that when $P$ is uniform, the eigenvectors $v_k$ converge to $\psi_k$ as $n \to \infty, \varepsilon \to 0$, and it is therefore natural that they accurately approximate only periodic functions $f_0 \in H^s([0,1])$. Although the zero-trace boundary condition is somewhat more restrictive than mere periodicity, the point is that some kind of boundary condition is inherently necessary in order to obtain sharp rates for spectral series estimators. For simplicity, we will stick to the specific boundary condition~\eqref{eqn:zero_trace}.

>>>>

\documentclass{article}
\usepackage{amsmath}
\usepackage{amsfonts, amsthm, amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\hypersetup{
	colorlinks=true,
	linkcolor=blue,
	citecolor=blue
}
\usepackage[parfill]{parskip}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{enumerate}
\usepackage[shortlabels]{enumitem}
\usepackage{fullpage}
\usepackage{mathtools}
\usepackage{tikz}
\usepackage{bm}
\usepackage{multirow}

\usepackage{natbib}
\renewcommand{\bibname}{References}
\renewcommand{\bibsection}{\subsubsection*{\bibname}}

@@

\newcommand{\emF}{\mathbb{F}_n}
\newcommand{\emG}{\mathbb{G}_n}
\newcommand{\emP}{\mathbb{P}_n}
\newcommand{\F}{\mathcal{F}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\Rd}{\Reals^d}
\newcommand{\RD}{\Reals^D}
\newcommand{\Nbb}{\mathbb{N}}

@@

Suppose we observe data $X_1,\ldots,X_n$ sampled independently from an unknown distribution $P$. As a replacement for $P$, we form a geometric graph $G$ over the observed data, with weighted edges $W_{ij}$ corresponding to proximity between samples $X_i$ and $X_j$. Geometric graphs encode information about $P$--- for instance, the number of connected components of its support, and the shape of each connected component---in an extremely general manner, and can be leveraged to conduct many different fundamental statistical tasks. These include clustering, manifold learning, semi-supervised learning, classification, and regression---both estimation and goodness-of-fit testing. Though much theoretical work has been done on the consistency of learning methods defined with respect to geometric graphs---and more recently, rates of convergence have been established for some methods---little is known so far about their optimality, even for classic statistical tasks.

In this paper we focus on the theoretical statistical properties of \textcolor{red}{graph-based regression}. We assume that in addition to the design points $X_1,\ldots,X_n$ one observes real-valued responses $Y_1,\ldots,Y_n$, and seeks to learn the unknown regression function $f_0(x) := E[Y|X = x]$. The methods we will study are based on \emph{Laplacian eigenmaps}, first introduced by \cite{belkin03a}, which takes the eigenvectors of the graph Laplacian $L$. The Laplacian $L$ is a difference operator, acting on functions $f: \{X_1,\ldots,X_n\} \to \Reals$ as follows,
\begin{equation}
\label{eqn:graph_laplacian}
Lf(X_i) = \sum_{j = 1}^{n} \bigl(f(X_i) - f(X_j)\bigr)W_{ij}. 
\end{equation}
The Laplacian $L$ can be seen as a discretization of the Laplace-Beltrami operator $\Delta_P$ \textcolor{red}{(...)}. The eigenvectors of $L$ form an orthonormal basis of $L^2(X)$, and are approximations to the eigenvectors of $\Delta_P$. Their corresponding eigenvalues are estimates of~\textcolor{red}{(...)}, and provide a notion of smoothness for each eigenvector---the smaller the eigenvalue, the smoother the corresponding eigenvector.

Laplacian eigenmaps is thus an example of a \emph{spectral series} estimator. A spectral series estimator is a special case of an orthogonal series estimator, one of the most classical methods of non-parametric regression. Traditionally, orthogonal series estimators are defined with respect to a fixed set of basis functions, orthogonalized in $L^2(Q)$ with respect to some pre-determined reference measure $Q$. Spectral series estimators take a particular orthogonal basis, the eigenfunctions of $\Delta_Q$. In contrast, eigenvectors of the graph Laplacian are data-dependent objects, and adapt to the geometry of the design distribution $P$ in a rich manner.

The theoretical properties of (classical) spectral series methods are various, and are in general extremely well-understood \textcolor{red}{(references)}. In particular, they are minimax rate optimal for nonparametric regression over H\"{o}lder and Sobolev spaces. However, in practice these methods suffer from some serious drawbacks. As a practical matter,  even finding the eigenfunctions of $\Delta_Q$ is in general non-trivial. The statistical optimality of such estimators also hinges on the reference measure $Q$ being equal to $P$, an unrealistic assumption when $P$ is unknown. If $Q \neq P$, the standard recommendation is to instead perform nonparametric least squares---this can be statistically rate-optimal, but is numerically unstable when the dimension $d$ is even moderately large.

\textcolor{red}{(TODO)}: Change the preceding paragraph.

Laplacian eigenmaps avoids these drawbacks \textcolor{red}{(...)} On the other hand its statistical properties are not as well understood \textcolor{red}{(...)}. Thus, there exists a theoretical gap.

\paragraph{Our contributions.} The primary contribution of our paper is to fill this theoretical gap, by answering the following question:

@@

Broadly speaking, we show that when the regression function $f_0$ is smooth, in the Sobolev sense of having weak derivatives bounded in $\Leb^2$-norm, Laplacian eigenmaps methods are statistically minimax optimal, for both estimation and goodness-of-fit testing. This statement holds for different relations between the dimension $d$ and number of derivatives $s$, depending on the problem (estimation or testing).
\begin{table}
	\begin{center}
		\begin{tabular}{p{.2\textwidth} | p{.14\textwidth} p{.12\textwidth} }
			Smoothness order & Flat Euclidean (Model~\ref{def:model_flat_euclidean}) & Manifold (Model~\ref{def:model_manifold}) \\
			\hline
			$s \leq 3$ & $\bm{n^{-2s/(2s + d)}}$ & $\bm{n^{-2s/(2s + m)}}$ \\
			$s > 3$  & $\bm{n^{-2s/(2s + d)}}$ & \textcolor{red}{(?)}
		\end{tabular}
	\end{center}
	\caption{Summary of estimation rates over Sobolev balls. Bold font marks minimax optimal rates. In each case, rates hold for all $d \in \mathbb{N}$ (under Model~\ref{def:model_flat_euclidean}), and for all $d, m \in \mathbb{N}$ (under Model~\ref{def:model_manifold}). Although we suppress it for simplicity, in all cases the dependence of the error rate on the radius $M$ of the Sobolev ball is also optimal, as long as $M \gtrsim n^{-1/2}$.}
	\label{tbl:estimation_rates}
\end{table}

\begin{table}
	\begin{center}
		\begin{tabular}{p{.175\textwidth} p{.175\textwidth} | p{.14\textwidth} p{.12\textwidth} }
			Smoothness order & Dimension & Flat Euclidean (Model~\ref{def:model_flat_euclidean}) & Manifold (Model~\ref{def:model_manifold}) \\
			\hline
			\multirow{2}{*}{$s = 1$} & $\dim(\mc{X}) < 4$ & $\bm{n^{-4s/(4s + d)}}$ & $\bm{n^{-4s/(4s + m)}}$ \\
			& $\dim(\mc{X}) \geq 4$ & $\bm{n^{-1/2}}$ & $\bm{n^{-1/2}}$ \\
			\hline
			\multirow{3}{*}{$s = 2$ or $3$} & $\dim(\mc{X}) \leq 4$  & $\bm{n^{-4s/(4s + d)}}$ & $\bm{n^{-4s/(4s + m)}}$ \\
			& $4 <\dim(\mc{X}) < 4s$  & \textcolor{red}{(?)}& \textcolor{red}{(?)} \\
			& $\dim(\mc{X}) \geq 4s$ & $\bm{n^{-1/2}}$ & $\bm{n^{-1/2}}$ \\
			\hline
			\multirow{3}{*}{$s > 3$} & $\dim(\mc{X}) \leq 4$ & $\bm{n^{-4s/(4s + d)}}$ & \textcolor{red}{(?)} \\
			& $4 < \dim(\mc{X}) < 4s$ & \textcolor{red}{(?)} & \textcolor{red}{(?)} \\
			& $\dim(\mc{X}) \geq 4s$ & $\bm{n^{-1/2}}$ & $\bm{n^{-1/2}}$ \\
		\end{tabular}
	\end{center}
	\caption{Summary of testing rates over Sobolev balls. Bold font marks minimax optimal rates. Rates when $d > 4s$ assume that $f_0 \in L^4(P,M)$. Although we suppress it for simplicity, in all cases the dependence of the error rate on the radius $M$ of the Sobolev ball is also optimal, as long as $M \gtrsim n^{-1/2}$.}
	\label{tbl:testing_rates}
\end{table}

\paragraph{Related Work.} 
\begin{itemize}
	\item \textcolor{red}{(TODO)} Summarize what is known on fixed graphs.
	\item \textcolor{red}{von Luxburg, Belkin, Slepcev} show that for a fixed index $k$, the eigenvector $v_k$ converges to the eigenfunction $u_k$, in the sense that \textcolor{red}{(...)}. \textcolor{red}{Shi, Burago, Garcia Trillos, Calder, Cheng} build on this, giving finite sample bounds, rates of convergence, and making statements uniform over $1 \leq k \leq k_{\max}$. This suggests that Laplacian eigenmaps may be good alternatives to classical spectral series methods when the latter cannot be implemented (e.g. because $P$ or $\mc{X}$ are unknown.) 
	\item There has been limited analysis of the estimator $\wh{f}$. \cite{zhou2011} consider a similar estimator, but in the semi-supervised setting where the number of unlabeled points grows to infinity. In this case the estimator reduces to a classical spectral series estimator, and there is no error incurred by approximating $\Delta_P$. \cite{lee2016} consider the \emph{diffusion maps} estimator---which uses the eigenvectors of a different normalization of the Laplacian $L$---in both the supervised and semi-supervised setups, but they derive suboptimal rates for the supervised problem. As far as we know, there has been no analysis of the test statistic $\wh{T}$ in the random design framework which we study.
	\item \textcolor{red}{(TODO)}: Alternatively, \cite{trillos2020} \textcolor{blue}{(Green 2021)} use the graph Laplacian operator to induce a penalty over functions $f: \{\bf X\} \to \Reals$. The \emph{Laplacian smoothing} estimator $\wh{f}_{\mathrm{LS}}$ is obtained by minimizing the sum of this penalty along with a data-fidelity term,
	\begin{equation*}
	\wh{f}_{\mathrm{LS}} := \|Y - f\|_n^2 + \lambda \dotp{L_{n,\varepsilon}f}{f}_n.
	\end{equation*}
	\textcolor{blue}{(Green 2021)} show that the resulting estimator is minimax optimal, but only for $s = 1$ and $d \leq 4$. In contrast, the Laplacian eigenmaps estimator is optimal for all $s$ and $d$. 
\end{itemize}

@@

\textcolor{red}{TODO}:

\begin{itemize}
	\item Define $\Leb^2(\mc{X}), \Leb^2(P)$ and $\Leb^2(P_n)$. Define $C^k(\mc{X})$, its norm, and its normed ball.
	\item Define asymptotic notation: $\ll$, $\lesssim$ and $\asymp$.
\end{itemize}

\section{Setup and Background}
\label{sec:setup_main_results}

In this section, we begin by giving a precise definition of our framework, and the Laplacian eigenmaps methods we study. We then review minimax rates for non-parametric regression over Sobolev spaces.

\subsection{Non-parametric regression with random design}
\label{sec:regression_laplacian_eigenmaps}

We will operate in the usual setting of non-parametric regression with random design, in which we observe independent random samples $(X_1,Y_1),\ldots,(X_n,Y_n)$. The design points $X_1,\ldots,X_n$ are sampled from a distribution $P$ with support $\mc{X} \subseteq \Rd$, and the responses follow
\begin{equation}
\label{eqn:model}
Y_i = f_0(X_i) + w_i,
\end{equation}
with regression function $f_0: \mc{X} \to \Reals$, and $w_i \sim N(0,1)$ independent Gaussian noise. For simplicity we will assume throughout that the noise has unit-variance, but all of our results extend in a straightforward manner to the case where the variance is equal to a known positive value.

We now formulate two models, which differ in the assumed nature of the support $\mc{X}$ of the design distribution $P$: the \emph{flat Euclidean} and \emph{manifold} models.

\begin{definition}[Flat Euclidean model]
	\label{def:model_flat_euclidean}
	The data $(X_1,Y_1),\ldots,(X_n,Y_n)$ are sampled according to~\eqref{eqn:model}. The support $\mc{X}$ of the design distribution $P$ is an open, connected, and bounded subset of $\Rd$, with Lipschitz boundary. The distribution $P$ admits a density $p$ with respect to the $d$-dimensional Lebesgue measure $\nu$, which is bounded away from $0$ and $\infty$,
	\begin{equation*}
	0 < p_{\min} \leq p(x) \leq p_{\max} < \infty, \quad \textrm{for all $x \in \mc{X}$.}
	\end{equation*}
\end{definition}

In the following, recall that the injectivity radius of a $m$-dimensional Riemannian manifold $\mc{X}$ is the maximum value of $\delta$ such that the exponential map $\exp_x: B_m(0,\delta) \subset T_x(\mc{X}) \to B_{\mc{X}}(x,\delta) \subset \mc{X}$ is a diffeomorphism for all $x \in \mc{X}$.
\begin{definition}[Manifold model]
	\label{def:model_manifold}
	The data $(X_1,Y_1),\ldots,(X_n,Y_n)$ are sampled according to~\eqref{eqn:model}. 
	The support $\mc{X}$ of the design distribution $P$ is a closed, connected, smooth and boundaryless Riemannanian manifold embedded in $\Rd$, of intrinsic dimension $1 \leq m < d$. The injectivity radius of $\mc{X}$ is upper bounded by a constant $i_0$. The design distribution $P$ admits a density $p$ with respect to the volume form $d\mu$ induced by the Riemannian structure, which is bounded away from $0$ and $\infty$,
	\begin{equation*}
	0 < p_{\min} \leq p(x) \leq p_{\max} < \infty, \quad \textrm{for all $x \in \mc{X}$.}
	\end{equation*}
\end{definition}

\subsection{Laplacian eigenmaps}
We now formally define the estimator and test statistic we study. Both are derived from eigenvectors of the neighborhood graph Laplacian operator. For a positive, symmetric kernel $\eta: [0,\infty) \to [0,\infty)$, and a bandwidth parameter $\varepsilon > 0$, the \emph{neighborhood graph Laplacian} operator $L_{n,\varepsilon}$ is
\begin{equation}
\label{eqn:neighborhood_graph_laplacian}
L_{n,\varepsilon}u(x) = \frac{1}{n\varepsilon^{2 + \mathrm{dim}(\mc{X})}} \sum_{i = 1}^{n} \bigl(u(x) - u(X_j)\bigr) \eta\biggl(\frac{\|x - X_j\|}{\varepsilon}\biggr)
\end{equation}
(Here $\mathrm{dim}(\mc{X})$ stands for the dimension of $\mc{X}$. It is equal to $d$ under the assumptions of Model~\ref{def:model_flat_euclidean}, and equal to $m$ under the assumptions of Model~\ref{def:model_manifold}. The pre-factor $(n\varepsilon^{2 + \mathrm{dim}(\mc{X})})^{-1}$ is purely for convenience in taking limits as $n \to \infty, \varepsilon \to 0$).

Although~\eqref{eqn:neighborhood_graph_laplacian} makes sense for any $x \in \mc{X}$ and $u: \mc{X} \to \Reals$, we will typically think of $x \in \{X_1,\ldots,X_n\}$ and $u = (u(X_1),\ldots,u(X_n)) \in \Leb^2(P_n)$), so that $L_{n,\varepsilon}$ is equivalently a matrix in $\Reals^{n \times n}$. Indeed, viewed as a matrix $L_{n,\varepsilon}$ is a symmetric and positive semi-definite, and can thus be diagonalized by its eigenpairs, solutions $(\lambda,v)$ to the eigenproblem
\begin{equation*}
\frac{1}{n}L_{n,\varepsilon}v = \lambda v, \quad \|v\|_n^2 = 1.
\end{equation*}
We will assume without loss of generality that each eigenvalue $\lambda$ of $L_{n,\varepsilon}$ has algebraic multiplicity $1$, and so we can index the eigenpairs $(\lambda_1,v_1),\ldots,(\lambda_n,v_n)$ in ascending order of eigenvalue, $0 = \lambda_1 < \ldots < \lambda_n$.

The Laplacian eigenmaps estimator $\wh{f}$ simply projects the response vector ${\bf Y}$ onto the first $K$ eigenvectors of $L_{n,\varepsilon}$: letting $V_K \in \Reals^{n \times K}$ be the matrix with columns $v_1,\ldots,v_K$, we have that
\begin{equation}
\label{eqn:laplacian_eigenmaps_estimator}
\wh{f} := \sum_{k = 1}^{K} \dotp{Y}{v_k}_{n} v_k = \frac{1}{n} V_K V_K^{\top} Y.
\end{equation} 
If $\wh{f}$ is a reasonable estimate of $f_0$, then the Laplacian eigenmaps test statistic
\begin{equation}
\label{eqn:laplacian_eigenmaps_test}
\wh{T} := \|\wh{f}\|_n^2
\end{equation}
is in turn a reasonable estimate of $\|f_0\|_{P}^2$, and can be used to distinguish whether or not $f_0 = 0$.

It may be helpful to comment briefly on the term ``Laplacian eigenmaps'', which we use a bit differently than is typical in the literature. Laplacian eigenmaps typically refers to an algorithm for embedding, which maps each design point $X_1,\ldots,X_n$ to $\Reals^K$ according to $X_i \mapsto (v_{1,i}, \ldots, v_{K,i})$. Viewing this mapping as a feature map, we can then interpret the estimator $\wh{f}$ as the least-squares solution to a linear regression problem with responses $Y_1,\ldots,Y_n$ and features $v_1,\ldots,v_K$. Often, the Laplacian eigenmaps embedding is viewed as a tool for dimensionality reduction, wherein it is implicitly assumed that $K \ll d$. We will neither explicitly nor implicitly take $K < d$; indeed, the embedding perspective is not particularly illuminating in what follows, and we do not henceforth make reference to it. Instead, we use ``Laplacian eigenmaps'' to directly refer to the estimator $\wh{f}$ or test statistic $\wh{T}$.

On a different note, it is worth explicitly pointing out that $\wh{f}$ is defined only in-sample, that is, only at the design points $X_1,\ldots,X_n$. There is no straightforward or agreed upon way to extend $\wh{f}$ out-of-sample, and we address the problem of out-of-sample estimation in some detail later in Section~\ref{sec:out_of_sample}.

\subsection{Sobolev Classes}
\label{sec:sobolev}
Before we give an overview of our main results, we will first pause to review the definition of Sobolev classes. We divide our discussion into two cases---the flat Euclidean case, and the manifold case, corresponding to Models~\ref{def:model_flat_euclidean} and Model~\ref{def:model_manifold}.

\paragraph{Flat Euclidean case.}
Take $\mc{X}$ to be an open, connected set with Lipschitz boundary, as in Model~\ref{def:model_flat_euclidean}. Let $Q$ be a measure supported on a subset of $\mc{X}$, that is absolutely continuous with respect to the full-dimensional Lebesgue measure $\nu$. Recall that for a given multiindex $\alpha \in \mathbb{N}^d$, a function $f$ is \emph{$\alpha$ weakly differentiable} if there exists some $h \in L^1(Q)$ such that
\begin{equation*}
\int_{\mc{X}} h g = (-1)^{|\alpha|} \int_{\mc{X}} f D^{\alpha}g, \quad \textrm{for every $g \in C_0^{\infty}(\mc{X})$.}
\end{equation*}
If such a function $h$ exists, it is the $\alpha$th weak partial derivative of $f$, and denoted by $D^{\alpha}f = h$. If $f$ is $\alpha$ weakly differentiable for all $|\alpha| = j$, then $d^jf := \sum_{|\alpha| = j}D^{\alpha}f$ is the sum of all weak partial derivatives of order $j$.

For an integer $s \geq 1$, the Sobolev class $H^s(Q)$ consists of all those functions $f \in \Leb^2(Q)$ which have weak derivatives of all orders $j = 1,\ldots,s$, all of which also belong to $\Leb^2(Q)$. The $j$th order semi-norm for $f \in H^s(Q)$ is $|f|_{H^j(Q)} := \|d^jf\|_{L^2(Q)}$. The corresponding norm
\begin{equation*}
\|f\|_{H^s(Q)}^2 := \|f\|_{\Leb^2(Q)}^2 + \sum_{j = 1}^{s} |f|_{H^j(Q)}^2,
\end{equation*}
which induces the ball
\begin{equation*}
H^s(Q;M) := \bigl\{f \in H^s(Q): \|f\|_{H^s(Q)} \leq M\bigr\}.
\end{equation*} 
In the special case where $Q$ is the uniform measure over $\mc{X}$, we use the notation $H^s(\mc{X}) := H^s(Q)$. Finally, we note that $H^s(Q)$ is the completion of $C^{(\infty)}(\mc{X})$ with respect to the $\|\cdot\|_{H^s(Q)}$ norm, so that $C^{\infty}(\mc{X})$ is dense in $H^s(\mc{X})$.

\paragraph{Manifold case.}

\paragraph{Boundary conditions.}
In the flat Euclidean case, when $s > 1$ we will assume that $f_0$ satisfies some boundary conditions. In particular, we will assume that $f_0$ belongs to the set of \emph{zero trace} Sobolev functions $H_0^{s}(P)$, defined as the closure of $C_0^s(\mc{X})$ with respect to $\|\cdot\|_{H^s(P)}$ norm. Of course, under Model~\ref{def:model_manifold} the domain $\mc{X}$ is without boundary, and so boundary conditions are irrelevant.
